---
title: "NZ Baby Names"
author: "Tim Hill"
date: "1 July 2016"
output: 
  html_document: 
    keep_md: yes
---

```{r, message=F, warning = F, echo = F}
library(dplyr)
library(tidyr)
library(ggplot2)
library(RColorBrewer)

babynames <- read.csv("NZnamedata.csv", header=T)
```


For the first couple of plots we'll look at three generations of my family - my parents, my sister and myself, and my two children.

```{r, echo = F}
girlsnames <- c("Elizabeth","Amelia","Ann")
boysnames <- c("Timothy", "Charlie","Roger")
selnames <- filter(babynames, (Name %in% girlsnames & Sex =="Girl") | 
                     (Name %in% boysnames & Sex =="Boy"))

```

Plot 1 shows the number of babies with the given names for the period 1954 - 2015

```{r, echo = F}
ggplot(data = selnames, aes(x = Year, y = No., colour = Name)) +
  geom_line(aes(colour = Name)) + 
  scale_color_brewer(palette = "Set2") +
  theme_bw() + 
  labs(title = "Plot 1 - Number of babies with selected names", y = "Babies named") +
  theme(legend.key = element_blank())
```

Plot 2 shows the ranking of the selected names (rank 1 = most popular name)

```{r, echo = F}
ggplot(data = selnames, aes(x = Year, y = Rank, colour = Name)) +
  geom_line(aes(colour = Name)) + 
  scale_color_brewer(palette = "Set2") +
  theme_bw() + 
  scale_y_reverse(breaks = seq(0,100,15)) + 
  labs(title = "Plot 2 - Ranking of selected names", y = "Name Rank") + 
  scale_x_continuous(breaks = seq(0, 2015, 10)) + 
  theme(legend.key = element_blank())
```

From the two plots above, you can probably guess what generation each of the names come from.


Now we'll dig into the data a bit deeper.  What are the most popular names in the period 1955 - 2015?

```{r, echo = F}
### Boys names
topboynames <- babynames %>% 
  filter(Sex == "Boy") %>% 
  group_by(Name) %>% 
  summarise(Number = sum(No.)) %>% 
  arrange(-Number) %>% 
  slice(1:10) %>% 
  mutate(Name = reorder(Name, -Number, sum)) 

ggplot(topboynames, aes(x = Name, y = Number)) + 
  geom_bar(stat="identity") +
  labs(title="Plot 3 - Most popular boys names 1955-2015", y = "Number of babies") +
  theme_bw()


### Girls names
topgirlnames <- babynames %>% 
  filter(Sex == "Girl") %>% 
  group_by(Name) %>% 
  summarise(Number = sum(No.)) %>% 
  arrange(-Number) %>% 
  slice(1:10) %>% 
  mutate(Name = reorder(Name, -Number, sum))

ggplot(topgirlnames, aes(x = Name, y = Number)) + 
  geom_bar(stat="identity") +
  labs(title="Plot 4 - Most popular girls names 1955-2015", y = "Number of babies") +
  theme_bw()

```


So what is the popularity of those names on a year by year basis?
For a cleaner plot, we'll only look at the top 6 for each sex.

```{r, echo = F}
### Top boys names

# Create vector of names
top6boysvec <- topboynames %>% select(Name) %>% slice(1:6) %>% droplevels() %>% .$Name

# Create df
top6boysdata <- babynames %>% filter(Name %in% top6boysvec & Sex == "Boy") %>% droplevels()

# Create plot  
ggplot(data = top6boysdata, aes(x = Year, y = Rank, colour = Name)) +
  geom_line(aes(colour = Name)) + 
  scale_color_brewer(palette = "Set2") +
  theme_bw() + 
  scale_y_reverse(breaks = c(1,10,20,30,40,50,60,70,80,90,100)) + 
  labs(title = "Plot 5 - Ranking of top boys names", y = "Name Rank") + 
  scale_x_continuous(breaks = seq(0, 2015, 10)) + 
  theme(legend.key = element_blank())

### Top girls names

# Create vector of required names
top6girlsvec <- topgirlnames %>% select(Name) %>% slice(1:6) %>% droplevels() %>% .$Name

# Create df 
top6girlsdata <- babynames %>% filter(Name %in% top6girlsvec & Sex == "Girl") %>% droplevels()

# Create plot  
ggplot(data = top6girlsdata, aes(x = Year, y = Rank, colour = Name)) +
  geom_line(aes(colour = Name)) + 
  scale_color_brewer(palette = "Set2") +
  theme_bw() + 
  scale_y_reverse(breaks = c(1,10,20,30,40,50,60,70,80,90,100)) + 
  labs(title = "Plot 6 - Ranking of girls names", y = "Name Rank") + 
  scale_x_continuous(breaks = seq(0, 2015, 10)) + 
  theme(legend.key = element_blank())

```

Generally speaking, the most popular names have shown a decline in recent years.

What about sudden changes in popularity?
To look at this, two new variables will be created.

- **difference** - this is the difference in the number of babies with that name from one year to the next.
- **diffrank** - the difference in ranking for that name from one year to the next

```{r, echo = F, cache = T}

### Need to add a variable with the annual difference - probably a cleaner way to do this
difference <- numeric(length = nrow(babynames))
diffrank <- numeric(length = nrow(babynames))

for(i in seq(1, nrow(babynames))) {
  currentyear <- babynames[i,1]
  currentname <- as.character(babynames[i,2])
  currentsex <- as.character(babynames[i,4])
  currentnum <- babynames[i,3]
  currentrank <- babynames[i,5]
  lastyear <- currentyear-1
  lastyear <- filter(babynames, Year == lastyear & Name == currentname & Sex == currentsex)
  lastnum <- ifelse(nrow(lastyear) == 0, 0, lastyear$No. )
  lastrank <- ifelse(nrow(lastyear) == 0, 101, lastyear$Rank )
  
  difference[i] <- (currentnum - lastnum)
  diffrank[i] <- (lastrank - currentrank)
}

babynames <- cbind(babynames, difference, diffrank)

#########



girlsdiffrank <- babynames %>% filter(Year != "1954" & Sex == "Girl") %>% arrange(-abs(diffrank)) %>% distinct(Name) %>% slice(1:6)
girlsdiffrank

plotdatagirls <- filter(babynames, Name %in% girlsdiffrank$Name)

ggplot(plotdatagirls, aes(x = Year, y = Rank, group = Name, colour = Name)) + 
  geom_line(aes(colour = Name)) +
  scale_color_brewer(palette = "Set2") +
  theme_bw() + 
  labs(title = "Plot 7 - Top 6 movers in girls names") +
  scale_y_reverse(breaks = c(1,10,20,30,40,50,60,70,80,90,100)) +
  theme(legend.key = element_blank())
```


```{r, echo = F, cache = T}

boysdiffrank <- babynames %>% filter(Year != "1954" & Sex == "Boy") %>% arrange(-abs(diffrank)) %>% distinct(Name) %>% slice(1:6)
boysdiffrank

plotdataboys <- filter(babynames, Name %in% boysdiffrank$Name)

ggplot(plotdataboys, aes(x = Year, y = Rank, group = Name, colour = Name)) + 
  geom_line(aes(colour = Name)) +
  scale_color_brewer(palette = "Set2") +
  theme_bw() + 
  labs(title = "Plot 8 - Top 6 movers in boys names") +
  scale_y_reverse(breaks = c(1,10,20,30,40,50,60,70,80,90,100)) +
  theme(legend.key = element_blank())


```

Some definite cultural events there - the name Brandon appearing from nowhere in 1992 due to 90210, and Jorja in 2003 - presumably thanks to Jorja Fox in CSI. 

The baby name data only contains details on the top 100 names for each year - what proportion of babies are included in this?  Are there many babies with names that don't make the top 100?

First, lets look at how many babies are included in the given data:

```{r, echo = F}
yeartotals <- babynames %>% 
  group_by(Sex, Year) %>% 
  summarise(Total.in.top.100 = sum(No.))

ggplot(yeartotals, aes(x = Year, y = Total.in.top.100, group = Sex)) +
  geom_line(aes(colour = Sex)) +
  scale_color_brewer(palette = "Set2") +
  theme_bw() +
  theme(legend.key = element_blank()) +
  labs(title = "Plot 9 - How many babies are included in the top 100 for each year?")

```

Unless the birth rate is dropping drastically, it appears that the top 100 names are accounting for a smaller and smaller proportion of new babies.


